---
title: "Image Processing and Spatial linear transformations"
author: "Manuel Mendes"
date: "Monday, June 22, 2015"
layout: post
category: R Mathematics Image Processing Matrix Operation
---

We can think of an **image** as a function, f, from  $$\pmb R^2 \rightarrow  R$$ (or a 2D signal):    

- f (x,y) gives the **intensity** at position (x,y)

Realistically, we expect the image only to be defined over a rectangle, with a finite range:    
f: [a,b]x[c,d] -> [0,1]

A color image is just three functions pasted together. We can write this as a “vector-valued” function:

$$\pmb {f(x,y)} = \bigg[ \begin{array}
{c}
& r(x,y) \cr \\
& g(x,y) \cr \\
& b(x,y) 
\end{array} \bigg] 
$$

### Image Processing

The package **EBImage** is an R package which provides general purpose functionality for the reading, writing, processing and analysis of images.


```{r}
# source("http://bioconductor.org/biocLite.R")
# biocLite()
# biocLite("EBImage")

# Reading Image
img <- readImage("images/lena_std.tif")

#str(img1)
dim(img)
display(img, method = "raster")


```

* Image Properties

Images are stored as multi-dimensional arrays containing the pixel intensities. All EBImage functions are also able to work with matrices and arrays. 

```{r}
print(img1)
```

* Adjusting Brightness

```{r}
img1 <- img + 0.2
img2 <- img - 0.3
display(img1, method = "raster")
display(img2, method = "raster")
```

* Adjusting Contrast

```{r}
img1 <- img * 0.5
img2 <- img * 2
display(img1, method = "raster")
display(img2, method = "raster")
```

* Gamma Correction

```{r}
img1 <- img ^ 4
img2 <- img ^ 0.9
img3 <- (img + 0.2) ^3
display(img1, method = "raster")
display(img2, method = "raster")
display(img3, method = "raster")
```

* Cropping Image

```{r}
img1 <-img[(1:400)+100, 1:400,]
display(img1, method = "raster")
```

### Spatial Transformation

Spatial image transformations are done with the functions resize, rotate, translate and the functions flip and flop to reflect images.

Next we show the functions flip, flop, rotate and translate:
```{r}
y <- flip(img)
display(y, title='flip(img)', method = "raster")

y = flop(img) 
display(y, title='flop(img)', method = "raster")

y <- rotate(img, 30) 
display(y, title='rotate(img, 30)', method = "raster")

y <- translate(img, c(120, -20)) 
display(y, title='translate(img, c(120, -20))', method = "raster")

```


#### All spatial transforms except flip and flop are based on the general affine transformation.

Linear transformations using the function **affine**:     

*  **Horizontal flip**

$${m} = \left[ 
\begin{array}{cc}
\ -1 & 0 \cr \\
\ 0 & 1 
\end{array}
\right] 
$$

$$
\begin{equation}
   Result = image * m
\end{equation}
$$


```{r}
m <- matrix(c(-1, 0, 0, 1, 512, 0), nrow=3,  ncol=2, byrow = TRUE)
m  # Horizontal flip
display(affine(img, m),  method = "raster")
```

*  **Horizontal shear**
$${m} = \left[ 
\begin{array}{cc}
1, 1/2 \cr \\
0, 1 
\end{array}
\right] 
$$


```{r}
m <- matrix(c(1, 1/2, 0, 1, 0, -100), nrow=3,  ncol=2, byrow = TRUE)
m  # horizontal shear  r = 1/2 
display(affine(img, m),  method = "raster")
```

*  **Rotation by π/6**
$${m} = \left[ 
\begin{array}{cc}
cos(pi/6), -sin(pi/6) \cr \\
sin(pi/6), cos(pi/6)
\end{array}
\right] 
$$

```{r}
m <- matrix(c(cos(pi/6), -sin(pi/6), sin(pi/6) , cos(pi/6), -100, 100), nrow=3,  ncol=2, byrow = TRUE)
m  # Rotation by π/6
display(affine(img, m),  method = "raster")
```

*  **Squeeze mapping with r=3/2**
$${m} = \left[ 
\begin{array}{cc}
3/2, 0 \cr \\
0, 2/3
\end{array}
\right] 
$$

```{r}
m <- matrix(c(3/2, 0, 0, 2/3, 0, 100), nrow=3,  ncol=2, byrow = TRUE)
m  # Squeeze mapping with r=3/2
display(affine(img, m),  method = "raster")
```

*  **Scaling by a factor of 3/2**
$${m} = \left[ 
\begin{array}{cc}
3/2, 0 \cr \\
0, 3/2
\end{array}
\right] 
$$

```{r}
m <- matrix(c(3/2, 0, 0, 3/2, -100, -100), nrow=3,  ncol=2, byrow = TRUE)
m  # Scaling by a factor of 3/2
display(affine(img, m),  method = "raster")
```

*  **Scaleing horizontally by a factor of 1/2** 
$${m} = \left[ 
\begin{array}{cc}
1/2, 0 \cr \\
0, 1
\end{array}
\right] 
$$

```{r}
m <- matrix(c(1/2, 0, 0, 1, 0, 0), nrow=3,  ncol=2, byrow = TRUE)
m  # scale a figure horizontally  r = 1/2 
display(affine(img, m),  method = "raster")
```


### References

* [http://www.iitg.ernet.in/scifac/qip/public_html/cd_cell/chapters/dig_image_processin.pdf](http://www.iitg.ernet.in/scifac/qip/public_html/cd_cell/chapters/dig_image_processin.pdf)
* [http://www.math.ksu.edu/research/i-center/UndergradScholars/manuscripts/FernandoRoman.pdf](http://www.math.ksu.edu/research/i-center/UndergradScholars/manuscripts/FernandoRoman.pdf)
* [http://www.statpower.net/Content/310/R%20Stuff/SampleMarkdown.html](http://www.statpower.net/Content/310/R%20Stuff/SampleMarkdown.html)
* [http://linear.ups.edu/html/section-LT.html](http://linear.ups.edu/html/section-LT.html)
* [http://www.r-bloggers.com/r-image-analysis-using-ebimage/](http://www.r-bloggers.com/r-image-analysis-using-ebimage/)
* [http://www.maa.org/external_archive/joma/Volume8/Kalman/Linear3.html](http://www.maa.org/external_archive/joma/Volume8/Kalman/Linear3.html)
* [https://en.wikipedia.org/wiki/Linear_map](https://en.wikipedia.org/wiki/Linear_map)
* [https://en.wikipedia.org/wiki/Matrix_%28mathematics%29](https://en.wikipedia.org/wiki/Matrix_%28mathematics%29)
* [https://en.wikipedia.org/wiki/Affine_transformation](https://en.wikipedia.org/wiki/Affine_transformation)









