---
title: "Map Visualization in R"
author: "Manuel Mendes"
date: "Wednesday, June 10, 2015"
layout: post
category: R
---

Here I tried to produce some map visualization in R.   

* First using data from [GADM database of Global Administrative Areas] (http://gadm.org/).           
* Second using the package RWorldMap, and            
* Third using the package ggmap that allows visualizations of spatial data on maps retrieved from Google Maps, OpenStreetMap, etc.      
* Fourth using the package RgoogleMaps allows you to plot data points on any kind  of map you can imagine  (terrain, satellite, hybrid).

# 1. Data from GADM database
## Getting the spatial country data

```{r, message=F, warning=F}
library(geosphere) # use: gcIntermediate
library(RgoogleMaps); library(ggmap); library(rworldmap)
library(sp); library(maptools); require(RColorBrewer)

# PRT_adm2.RData download from http://www.gadm.org/
# load("data/PRT_adm2.RData") # Creates and stores in memory an object called ´gadm´
# or:
# con <- url("http://biogeo.ucdavis.edu/data/gadm2/R/PRT_adm1.RData")
# print(load(con))
# close(con)
# or:
# library(raster)
# port1 <- getData('GADM', country='PRT', level=1)

load("data/PRT_adm1.RData")
port1 <- get("gadm")

```

## Plot the map
We can use the variable NAME_1 to plot the map:

```{r}

# Convert the encoding to UTF-8 in order to avoid the problems with 'accents'.
# port1$NAME_1 <- as.factor(iconv(as.character(port1$NAME_1), , "UTF-8"))
port1$NAME_1 <- as.factor(as.character(port1$NAME_1))

spplot(port1, "NAME_1", 
       col.regions = colorRampPalette(brewer.pal(12, "Set3"))(18), 
       col = "white",
       #cex=0.4,
       xlim = range(-10,-6), ylim = range(36.9,42.2), asp = 1.0
       #scales=list(draw=TRUE)
       )  # Plot the 'NAME_1' form the 'port1' object.

```

And next we created another variable called rainfall in the data.frame, we store random values in that variable:

```{r}
set.seed(333)
port1$rainfall<-rnorm(length(port1$NAME_1),mean=50,sd=15) #random rainfall value allocation

spplot(port1,"rainfall",col.regions = rev(terrain.colors(port1$rainfall)),
       scales=list(draw=TRUE),
       main="Rainfall (simulated) in Portugal")

```

# 2. RWorldMap
Rworldmap is a package for visualising global data, referenced by country. It provides maps as spatial polygons.

```{r}
newmap <- getMap(resolution = "low")
plot(newmap)

```

By Changing the xlim and ylim arguments of the plot function we can limit the display to just Europe.

```{r}
plot(newmap, xlim = c(-20, 59),  ylim = c(35, 71),  asp = 1 )

```

## [Geocoding] (https://en.wikipedia.org/wiki/Geocoding)
 
The geocode function from the ggmap package finds the coordinates of a location using Google Maps. Thus, finding the coordinates of the [Extreme points of Europe] (https://en.wikipedia.org/wiki/Extreme_points_of_Europe) can be done by: 

```{r}
europe.limits <- geocode(c("CapeFligely,RudolfIsland,Franz Josef Land,Russia", 
                           "Gavdos,Greece", "Faja Grande,Azores",
                           "SevernyIsland,Novaya Zemlya,Russia"))
europe.limits

```

So we can display only the europe as follow, by modifying the xlim and ylim arguments:

```{r}
plot(newmap, xlim = range(europe.limits$lon),  ylim = range(europe.limits$lat), asp = 1)

```

# 3. ggmap   
* Fetching a Map

```{r}
# geocodes
geoCodes <- geocode("Portela,2230")
geoCodes
#         lon      lat
# 1 -8.250595 39.61002
```

```{r, message=F, warning=F}
ggmap(
    get_googlemap(
        center=c(geoCodes$lon,geoCodes$lat), #Long/lat of centre
        zoom=14, 
        maptype='satellite', #also hybrid/terrain/roadmap
        scale = 2), #resolution scaling, 1 (low) or 2 (high)
    size = c(600, 600), #size of the image to grab
    extent='device', #can also be "normal" etc
    darken = 0) #you can dim the map when plotting on top

#ggsave ("images/map1.png", dpi = 200) #this saves the output to a file

```

* Plotting on a Map

You can plot any [x,y, +/- z] information you’d like on top of a ggmap, so long as x and y correspond to longitudes and latitudes within the bounds of the map you have fetched. To plot on top of the map you must first make your map a variable and add a geom layer to it.

```{r, message=F, warning=F}
#Generate some data
#         lat        lon
# Colmeal 39.609025, -8.245660
# Portela 39.610447, -8.248321
# Cabeça_Ruiva 39.606380, -8.259393
# Ilha_do_Lombo 39.609124, -8.271237
long = c(-8.245660, -8.248321, -8.259393, -8.271237)
lat = c(39.609025, 39.610447, 39.606380, 39.609124)
who = c("Colmeal", "Portela", "Cabeça Ruiva", "Ilha do Lombo")
data = data.frame (long, lat, who)

map = ggmap(
    get_googlemap(
		center=c(geoCodes$lon,geoCodes$lat), 
		zoom=14, 
		maptype='hybrid', 
		scale = 2), 
		size = c(600, 600),
		extent='normal', 
		darken = 0)

map + geom_point (
		data = data,
		aes (
			x = long, 
			y = lat, 
			fill = factor (who)
			), 
		pch = 21, 
		colour = "white", 
		size = 6
		) +

	scale_fill_brewer (palette = "Set1", name = "Local") +

	#for more info on these type ?theme()	
	theme ( 
		legend.position = c(0.05, 0.05), # put the legend INSIDE the plot area
		legend.justification = c(0, 0),
		legend.background = element_rect(colour = F, fill = "white"),
		legend.key = element_rect (fill = F, colour = F),
		panel.grid.major = element_blank (), # remove major grid
		panel.grid.minor = element_blank (),  # remove minor grid
		axis.text = element_blank (), 
		axis.title = element_blank (),
		axis.ticks = element_blank ()
		) 

#ggsave (""images/map2.png"", dpi = 200)
```

## 4. RgoogleMaps
'RgoogleMaps' allows you to plot data points on any kind  of map you can imagine  (terrain, satellite, hybrid).

```{r, message=F, warning=F}

lat <- c(37,42) #define our map's ylim
lon <- c(-9,-6) #define our map's xlim
lat <- c(37,42) #define our map's ylim
lon <- c(-12,-6) #define our map's xlim
center = c(mean(lat), mean(lon))  #tell what point to center on
zoom <- 7  #zoom: 1 = furthest out (entire globe), larger numbers = closer in

# get maps from Google
terrmap <- GetMap(center=center, zoom=zoom, maptype= "terrain", destfile = "terrain.png") 
# visual options: 
#   maptype = c("roadmap", "mobile", "satellite", "terrain", "hybrid", "mapmaker-roadmap", "mapmaker-hybrid")
PlotOnStaticMap(terrmap)

# Sard <- geocode("Sardoal, PT") # find coordinates
# Sard

# using maps to plot routes
rt1 = route(from = "Lisbon", to = "Castelo Branco", mode = "driving")
rt2 = route(from = "Lisbon", to = "Sardoal", mode = "driving")
rt3 = route(from = "Abrantes", to = "Castelo Branco", mode = "driving")

PortugalMap <- qmap("Portugal", zoom = 8, color = "bw")

PortugalMap + geom_leg(aes(x = startLon, y = startLat, xend = endLon, yend = endLat),
                       color = "blue", data = rt1) +
    geom_leg(aes(x = startLon, y = startLat, xend = endLon, yend = endLat),
                       color = "black", data = rt2) +
    geom_leg(aes(x = startLon, y = startLat, xend = endLon, yend = endLat),
                       color = "red", data = rt3)



```


### References

* [OpenStreeMap] (http://download.geofabrik.de/)    
* [http://www.molecularecologist.com/2012/09/making-maps-with-r/] (http://www.molecularecologist.com/2012/09/making-maps-with-r/)        
* [http://allthingsr.blogspot.pt/2012/03/geocode-and-reverse-geocode-your-data.html] (http://allthingsr.blogspot.pt/2012/03/geocode-and-reverse-geocode-your-data.html)       
* [http://www.r-bloggers.com/13-mapping-in-r-representing-geospatial-data-together-with-ggplot/] (http://www.r-bloggers.com/13-mapping-in-r-representing-geospatial-data-together-with-ggplot/       )       
* [https://dl.dropboxusercontent.com/u/24648660/ggmap%20useR%202012.pdf] (https://dl.dropboxusercontent.com/u/24648660/ggmap%20useR%202012.pdf)       
* [https://wilkinsondarren.wordpress.com/2013/02/01/mapping-in-r-representing-geospatial-data-together-with-ggplot/] (https://wilkinsondarren.wordpress.com/2013/02/01/mapping-in-r-representing-geospatial-data-together-with-ggplot/)     
* [https://www.google.pt/maps/@39.6100171,-8.2505952,15z?hl=pt-PT] (https://www.google.pt/maps/@39.6100171,-8.2505952,15z?hl=pt-PT)          
* [http://www.di.fc.ul.pt/~jpn/r/maps/index.html] (http://www.di.fc.ul.pt/~jpn/r/maps/index.html)   
* [http://www.milanor.net/blog/?p=594] (http://www.milanor.net/blog/?p=594)     
* [http://pakillo.github.io/R-GIS-tutorial/] (http://pakillo.github.io/R-GIS-tutorial/)    
    
